---
title: "Disease Surveillance & Outbreak Analytics with R"
subtitle: "Day 1: Importing, Exploring & Cleaning Outbreak Data"
author: "Your Name | Your Organization"
date: "Saturday, 11 July 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    footer: "Disease Surveillance with R — Day 1"
    logo: ""
    incremental: false
    code-line-numbers: true
    highlight-style: github
    toc: false
execute:
  echo: true
  eval: false
---

## Welcome! {background-color="#1f3b57"}

::: {style="color:white;"}
**Disease Surveillance and Outbreak Analytics with R**

Free launch course — hands-on, practical, beginner-friendly

10:00 AM – 12:30 PM WAT, via Zoom
:::

---

## Housekeeping

- Session is **hands-on** — please have R & RStudio open and ready
- Slides + code + data: link shared in the chat / follow-up email
- Ask questions in the chat as we go; we'll pause for Q&A regularly
- Camera/mic optional; this is a working session

---

## Who is this course for?

- Public health professionals
- Epidemiology workers and students
- Disease surveillance officers
- M&E officers and health data analysts
- Researchers and postgraduate students
- NGO and health program staff

*No prior R experience assumed — we go step by step.*

---

## What you will take away today

By the end of Day 1, you will be able to:

1. Import a line-list dataset into R
2. Do a first exploration of an unfamiliar dataset
3. Clean messy real-world surveillance data:
   - inconsistent text (sex, ward, outcome)
   - messy/impossible ages
   - dates in multiple formats
   - duplicate records

---

## Today's scenario

> A suspected **Acute Watery Diarrhoea (AWD)** outbreak has been reported
> in **Ogun North LGA**. A raw line list has just arrived from the field
> data collection tool. Before anyone can analyze it — it needs cleaning.

We will use this *simulated* line list (`awd_linelist_raw.csv`)
throughout the course.

---

## Course roadmap

| Day | Focus |
|---|---|
| **Day 1 (today)** | R basics, import, explore, clean a line list |
| **Day 2 (tomorrow)** | Descriptive tables, epidemic curves, outbreak measures |

Course content adapted from **The Epidemiologist R Handbook**
([epirhandbook.com](https://www.epirhandbook.com/en/)) — a free, open
resource you should bookmark today.

---

# Part 1: R & RStudio Basics {background-color="#2e5e4e"}

::: {style="color:white;"}
(10:00 – 10:25)
:::

---

## What is R? What is RStudio?

- **R**: the programming language/engine that does the calculations
- **RStudio**: the friendly "cockpit" (IDE) we use to write and run R code
- **Packages**: add-on toolkits (e.g. `tidyverse` for data wrangling,
  `ggplot2` for graphs) — install once, load every session

---

## The RStudio layout

- **Source** (top-left): where you write scripts (`.R` files)
- **Console** (bottom-left): where code actually runs
- **Environment** (top-right): objects you've created (data, values)
- **Files/Plots/Packages/Help** (bottom-right)

*Live demo: tour of RStudio*

---

## R Projects — start here every time

```r
# File > New Project > New Directory > New Project
# Keep your script, data/, and outputs together in ONE folder
```

Using an **R Project** means your file paths (`"data/file.csv"`)
work no matter whose computer you're on.

---

## Core building blocks

```{r}
# assigning values
n_cases <- 353

# a vector
ages <- c(4, 17, 29, 5, 61)

# a function
mean(ages)

# a data frame (like a spreadsheet)
df <- data.frame(id = 1:3, age = c(4, 17, 29))
```

---

## The pipe `%>%` (or `|>`)

Reads as **"then"** — chains steps together instead of nesting functions.

```{r}
# Without pipe (hard to read)
round(mean(ages), 1)

# With pipe (reads top-to-bottom)
ages %>%
  mean() %>%
  round(1)
```

We will use the pipe constantly today.

---

## Key packages for this course

```{r}
install.packages(c("tidyverse", "janitor", "lubridate",
                    "here", "skimr"))

library(tidyverse)  # dplyr, ggplot2, readr, stringr, tidyr
library(janitor)    # clean_names(), tabyl(), get_dupes()
library(lubridate)  # dates made easy
library(skimr)      # quick data summaries
```

---

# Part 2: Importing & Exploring Data {background-color="#2e5e4e"}

::: {style="color:white;"}
(10:25 – 10:50) — Epi R Handbook Ch. 7 & 8
:::

---

## Importing your line list

```{r}
library(tidyverse)

linelist_raw <- read_csv("data/awd_linelist_raw.csv")
```

`read_csv()` (from `readr`) is faster and more predictable than
base R's `read.csv()`, and prints a helpful column-type summary.

---

## First look at any new dataset

```{r}
dim(linelist_raw)        # rows x columns
glimpse(linelist_raw)    # compact view of every column
head(linelist_raw)       # first 6 rows
```

```{r}
library(skimr)
skim(linelist_raw)       # full summary: missingness, ranges, etc.
```

::: {.callout-note}
**Habit to build:** never analyze a dataset you haven't looked at first.
:::

---

## 🖥️ Hands-on Exercise 1 (10 min)

1. Import `data/awd_linelist_raw.csv`
2. Run `glimpse()` and `skim()`
3. In the chat, tell us: how many rows? Which column has the most
   missing values?

---

# Part 3: Cleaning the Line List {background-color="#2e5e4e"}

::: {style="color:white;"}
(10:50 – 11:55) — Epi R Handbook Ch. 8, 9, 10, 15
:::

---

## Why real surveillance data is messy

- Multiple data collectors → inconsistent spelling/casing
- Manual date entry → mixed formats
- Weak network / re-submission → duplicate records
- Skipped fields → missing values
- Fat-finger errors → impossible values (age = 150)

**Cleaning is not optional — it *is* the analysis.**

---

## Step 1: Clean column names

```{r}
library(janitor)

linelist <- linelist_raw %>%
  clean_names()

names(linelist)
```

`clean_names()` turns `"Sex "`, `" Ward"`, `"date of report"` into
tidy `snake_case` names: `sex`, `ward`, `date_of_report`.

---

## Step 2: Find & remove duplicates

```{r}
get_dupes(linelist)              # inspect exact duplicates
linelist <- linelist %>% distinct()
```

**Near-duplicates** (e.g. `"AWD-0043"` vs `"AWD-0043b"`) need a rule:

```{r}
linelist <- linelist %>%
  mutate(case_id = str_remove(case_id, "[a-z]$")) %>%
  distinct(case_id, .keep_all = TRUE)
```

---

## Step 3: Clean categorical variables

```{r}
tabyl(linelist, sex)   # ALWAYS inspect before cleaning
```

```{r}
linelist <- linelist %>%
  mutate(sex = case_when(
    str_to_upper(str_trim(sex)) %in% c("M", "MALE")   ~ "Male",
    str_to_upper(str_trim(sex)) %in% c("F", "FEMALE") ~ "Female",
    TRUE ~ NA_character_
  ))
```

`case_when()` is your best friend for recoding messy categories.

---

## Step 3 (continued): ward names

```{r}
tabyl(linelist, ward)
```

```{r}
linelist <- linelist %>%
  mutate(
    ward = str_trim(ward),
    ward = str_to_title(ward),
    ward = recode(ward,
                  "Iwo Rd"  = "Iwo Road",
                  "Ring Rd" = "Ring Road")
  )
```

**Pattern:** trim whitespace → standardize case → fix known typos.

---

## 🖥️ Hands-on Exercise 2 (15 min)

Using the same pattern, clean:

1. `hospitalized` → should only contain `"Yes"` / `"No"`
2. `outcome` → should only contain `"Recovered"` / `"Died"` / `NA`

*Hint: use `tabyl()` first to see what you're dealing with, then
`case_when()`.*

(Solutions in `R/day1_import_and_cleaning.R`)

---

## Step 4: Clean numeric variables (age)

```{r}
summary(linelist$age)     # spot the impossible max!
```

```{r}
linelist <- linelist %>%
  mutate(age = if_else(age < 0 | age > 100, NA_real_, age))
```

Then create age groups for later analysis:

```{r}
linelist <- linelist %>%
  mutate(age_group = cut(age,
    breaks = c(0, 5, 15, 30, 45, 60, Inf),
    labels = c("0-4","5-14","15-29","30-44","45-59","60+"),
    right = FALSE))
```

---

## Step 5: Clean dates — the tricky part

Our data has THREE date formats mixed together:

- `2026-05-03`  (ISO: year-month-day)
- `03/05/2026`  (day/month/year)
- `03-May-2026` (day-month abbreviation-year)

```{r}
library(lubridate)

linelist <- linelist %>%
  mutate(date_onset = parse_date_time(
    date_onset, orders = c("ymd", "dmy", "mdy"), quiet = TRUE
  ) %>% as_date())
```

---

## ⚠️ A word of caution on dates

`11/05/2026` is genuinely ambiguous: is that 11 May or 5 November?

- Always **ask your data source** which format was used
- Where possible, request ISO format (`YYYY-MM-DD`) at collection stage
- After parsing, always **sanity-check the range** of resulting dates

```{r}
range(linelist$date_onset, na.rm = TRUE)
```

---

## Step 5 (continued): validate & fix dates

```{r}
linelist <- linelist %>%
  filter(!is.na(date_onset),
         date_onset >= as_date("2026-01-01"),
         date_onset <= as_date("2026-07-04")) %>%
  mutate(
    date_of_report = if_else(
      date_of_report < date_onset | is.na(date_of_report),
      date_onset, date_of_report),
    reporting_delay = as.numeric(date_of_report - date_onset)
  )
```

---

## 🖥️ Hands-on Exercise 3 (15 min)

1. Apply the date-cleaning code to `date_of_report`
2. Calculate `reporting_delay` (in days)
3. Find: what is the **maximum** reporting delay in the dataset?
   Is that concerning for surveillance timeliness?

---

## Save your cleaned dataset

```{r}
write_csv(linelist, "data/awd_linelist_clean_inclass.csv")
```

**Bring this file to Day 2!** We'll use it to build epidemic curves
and calculate outbreak measures.

---

## Recap: Day 1

✅ R & RStudio basics, the pipe
✅ Importing and first-look exploration
✅ Cleaning: column names, duplicates, categories, numbers, dates
✅ A clean, analysis-ready line list

**Tomorrow:** turn this clean data into epidemic curves, tables, and
outbreak measures (attack rate, CFR).

---

## Questions? {background-color="#1f3b57"}

::: {style="color:white;"}
See you tomorrow!

Sunday, 12 July 2026, 1:30 – 3:30 PM WAT
:::
