---
title: "Disease Surveillance & Outbreak Analytics with R"
subtitle: "Day 2: Epidemic Curves, Tables & Outbreak Measures"
author: "Adejumo Ridwan Suleiman | Business Data Laboratory"
date: "Sunday, 12 July 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    footer: "Disease Surveillance with R — Day 2"
    code-line-numbers: true
    highlight-style: github
    toc: false
execute:
  echo: true
  eval: false
---


## Quick recap of Day 1

- We imported a messy AWD line list
- Cleaned column names, duplicates, categories, ages, dates
- Saved: `data/awd_linelist_clean_inclass.csv`

Today we answer the questions a surveillance team actually needs:
**Who is affected? Where? When? How severe?**

---

## What you will take away today

1. Descriptive tables (age, sex, ward breakdowns)
2. Epidemic curves — the single most important outbreak graphic
3. Key outbreak measures: **attack rate**, **case fatality ratio**
4. A simple trend line (moving average)
5. A one-page surveillance summary

---

## Setup: load your data

```{r}
library(tidyverse)
library(janitor)
library(lubridate)

linelist <- read_csv("data/awd_linelist_clean_inclass.csv")
population <- read_csv("data/ward_population.csv")

# didn't finish Day 1? use the following data instead:
# linelist <- read_csv("data/awd_linelist_clean.csv")
```

---

# Part 1: Descriptive Tables {background-color="#2e5e4e"}

---

## `tabyl()` — quick frequency tables

```{r}
tabyl(linelist, sex) %>%
  adorn_pct_formatting(digits = 1)
```

```{r}
tabyl(linelist, ward) %>%
  arrange(desc(n))
```

Simple, readable, presentation-ready with one extra line of code.

---

## Cross-tabulations

```{r}
linelist %>%
  tabyl(age_group, sex) %>%
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns()
```

Answers: *"What proportion of each age group is male vs female?"*

---

# Part 2: Epidemic Curves {background-color="#2e5e4e"}


## Why the epicurve matters

The epidemic curve is usually the **first graphic** shown in any
outbreak report or sitrep. It tells you at a glance:

- Is the outbreak growing, peaking, or declining?
- Is transmission a single point-source or ongoing/propagated?
- Roughly when did exposure/transmission occur?

---

## A basic epicurve

```{r}
ggplot(linelist, aes(x = date_onset)) +
  geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") +
  labs(title = "Epidemic Curve: Suspected AWD Cases",
       x = "Date of onset", y = "Number of cases") +
  theme_minimal()
```


---

## A weekly epicurve, by outcome

```{r}
linelist <- linelist %>%
  mutate(epiweek = floor_date(date_onset, "week", week_start = 1))

ggplot(linelist, aes(x = epiweek, fill = outcome)) +
  geom_bar(colour = "white") +
  scale_fill_manual(values = c("Recovered" = "seagreen",
                                "Died" = "firebrick")) +
  labs(title = "Weekly Cases by Outcome",
       x = "Week of onset", y = "Cases") +
  theme_minimal()
```

---

## Faceting: one panel per ward

```{r}
ggplot(linelist, aes(x = epiweek)) +
  geom_bar(fill = "darkorange") +
  facet_wrap(~ ward) +
  labs(title = "Weekly Cases by Ward") +
  theme_minimal()
```

Facets let you spot **where transmission is still active** — very
useful for targeting a response.

---

# Part 3: Key Outbreak Measures {background-color="#2e5e4e"}

---

## Attack Rate

$$
\text{Attack Rate} = \frac{\text{New cases in a period}}{\text{Population at risk}} \times 10{,}000
$$

```{r}
attack_rate <- linelist %>%
  count(ward, name = "cases") %>%
  left_join(population, by = "ward") %>%
  mutate(attack_rate_per_10000 =
           round(cases / population * 10000, 1)) %>%
  arrange(desc(attack_rate_per_10000))

attack_rate
```

---

## Visualizing attack rate by ward

```{r}
ggplot(attack_rate,
       aes(x = reorder(ward, attack_rate_per_10000),
           y = attack_rate_per_10000)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  labs(title = "Attack Rate by Ward",
       x = NULL, y = "Cases per 10,000 population") +
  theme_minimal()
```

This immediately tells the response team **where to focus**.

---

## Case Fatality Ratio (CFR)

$$
\text{CFR} = \frac{\text{Deaths}}{\text{Cases with known outcome}} \times 100
$$

```{r}
linelist %>%
  filter(!is.na(outcome)) %>%
  summarise(
    n = n(),
    deaths = sum(outcome == "Died"),
    cfr_percent = round(deaths / n * 100, 2)
  )
```

::: {.callout-note}
Always report the **denominator** (cases with *known* outcome) —
don't silently drop missing outcomes without saying so.
:::

---

## Other quick indicators

```{r}
# % hospitalized
linelist %>%
  filter(!is.na(hospitalized)) %>%
  summarise(pct_hosp = round(mean(hospitalized == "Yes") * 100, 1))

# reporting timeliness
linelist %>%
  summarise(median_delay = median(reporting_delay, na.rm = TRUE),
            max_delay = max(reporting_delay, na.rm = TRUE))
```

Timeliness indicators matter as much as clinical ones — a slow
system can't respond to a fast outbreak.

---


## A simple trend: 7-day moving average

```{r}
daily_cases <- linelist %>%
  count(date_onset, name = "cases") %>%
  complete(date_onset = seq.Date(min(date_onset), max(date_onset), "day"),
           fill = list(cases = 0)) %>%
  mutate(ma7 = zoo::rollmean(cases, 7, fill = NA, align = "right"))

ggplot(daily_cases, aes(date_onset)) +
  geom_col(aes(y = cases), fill = "grey80") +
  geom_line(aes(y = ma7), colour = "darkblue", linewidth = 1) +
  theme_minimal()
```

Smooths day-to-day noise so you can see the underlying trend.

---

# Part 4: Putting It Together {background-color="#2e5e4e"}

---

## A one-page surveillance summary

```{r}
tibble(
  Indicator = c("Total cases", "Wards affected",
                "Overall attack rate /10,000", "CFR (%)",
                "% Hospitalized", "Median reporting delay (days)"),
  Value = c(nrow(linelist), n_distinct(linelist$ward),
            17.4, 2.1, 34.8, 2)   # <- replace with your calculated values
)
```

This table + epicurve + attack-rate chart = the core of any
outbreak **situation report (sitrep)**.

---

## Beyond today: automating reports

What we did manually today (import → clean → analyze → visualize)
can be **automated** with R Markdown / Quarto reports that
regenerate themselves every time new data arrives.

*(Covered in depth in our full training program.)*

---

## Recap: What we covered

✅ Day 1: Import, explore, clean a messy line list

✅ Day 2: Descriptive tables, epidemic curves, outbreak measures,
a simple trend, and a summary output

**You now have a repeatable workflow you can apply to real
surveillance data in your own work.**

---

## Thank you! {background-color="#1f3b57"}

::: {style="color:white;"}
Questions & discussion

Details on the full training program coming soon
:::
